ODGI: understanding pangenome graphs

### 2021-09-30 manuscript/main.tex
discussion draft, metrics text

Thanks to advances in sequencing technologies, new full-length genome assemblies are produced at a high rate, including efforst from the Human Pangenome Reference Consortium\footnote{https://humanpangenome.org/ (accessed Oct 2021)} (HPRC) and Telomere-to-Telomere (T2T) projects \citep{Miga:2020, Logsdon_2021, Nurk_2021}, from the T2T consortium\footnote{https://sites.google.com/ucsc.edu/t2tworkinggroup/home (accessed Oct 2021)}. This amount of data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full genetic variation across a population, but graph-based data structures present computational overheads: in addition to the sequences, they must also represent the graph topology. Several genome data structures were developed to address these challenges \citep{Eizenga_2020_BX}. Memory-efficient representations do not allow dynamic updates to the graph, such as adding new genomes or simplifying the graph topology. Furthermore, dynamic updates must perform efficiently on the complex graphs generated from whole-genome alignment of vertebrates. Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs that present paths with very high-depth nodes. The node depth is defined as the number of times in which the node is crossed by all the paths embedded in the graph. High depth increases the complexity of the operations performed on the paths and the nodes they cross.

@@

% simon
% chr8 longs

Applying odgi stats, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length. ODGI also offers more advanced tools for interrogating the graphs. odgi bin summarizes the path information into bins of specified size, enabling a summarized view of gigabase scale graphs. odgi depth returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to sequences with highly identical repeats. Complex motifs can be also detected with odgi degree, which returns the node degree as defined by query criteria. High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the process as well.
Solves the problem of understanding the overall structure, size, and features of the graph and its sequences.

Figure: small graph with all the stats shown. A screenshot of MultiQC, visualize summary per bin somehow, R visualizations of degree and depth HTT exon 1 as tiny example graph

@@

@@

>>>>

Thanks to advances in sequencing technologies, new full-length genome assemblies are produced at a high rate, including efforst from the Human Pangenome Reference Consortium\footnote{\url{https://humanpangenome.org/} (accessed Oct 2021)} (HPRC) and Telomere-to-Telomere (T2T) projects \citep{Miga:2020, Logsdon_2021, Nurk_2021}, from the T2T consortium\footnote{\url{https://sites.google.com/ucsc.edu/t2tworkinggroup/home} (accessed Oct 2021)}. This amount of data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full genetic variation across a population, but graph-based data structures present computational overheads: in addition to the sequences, they must also represent the graph topology. Several genome data structures were developed to address these challenges \citep{Eizenga_2020_BX}. Memory-efficient representations do not allow dynamic updates to the graph, such as adding new genomes or simplifying the graph topology. Furthermore, dynamic updates must perform efficiently on the complex graphs generated from whole-genome alignment of vertebrates. Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs that present paths with very high-depth nodes. The node depth is defined as the number of times in which the node is crossed by all the paths embedded in the graph. High depth increases the complexity of the operations performed on the paths and the nodes they cross.

@@

@@

% simon
% HTT exon1 
% Only chr8 longs if I can't produce someting nice

% Solves the problem of understanding the overall structure, size, and features of the graph and its sequences.

Graphs statistics provide alternative ways to gain insight into pangenomes complexity revealing the overall structure, size, and features of a graph and its sequences.

Applying \texttt{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length. MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs. ODGI also offers more advanced tools for the interrogation of the graphs. \texttt{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to collapsed sequences with highly identical repeats. Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for validation.

% Figure: small graph with all the stats shown. A screenshot of MultiQC, visualize summary per bin somehow, R visualizations of degree and depth HTT exon 1 as tiny example graph

@@

Pangenome graphs stand to become a ubiquitous tool in genomics \citep{Eizenga_2020}. With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale. Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs. ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs. Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder\footnote{\url{https://github.com/pangenome/pggb} (accessed Oct 2021)} (PGGB) or nf-core/pangenome\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}. Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.
